---
title: "EDUC 645 Syllabus"
site: distill::distill_website
output:
  html_document:
    toc: TRUE
    toc_depth: 3
    toc_float: TRUE
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# General Course Information


# Course Overview


# Student Learning Outcomes


# Textbooks and Reading Materials
We will primarily be referring to chapters in [Beyond Multiple Linear Regression ](https://bookdown.org/roback/bookdown-BeyondMLR/) (**BMLR**) by Paul Roback and Julie Legler. This textbook is available for free online. You may choose to purchase a paper copy if you wish, but it is not required. 


## R and RStudio
Students must have the latest version of R, which can be downloaded [here](https://ftp.osuosl.org/pub/cran/). It is strongly recommended that students also download the RStudio GUI, available [here](https://www.rstudio.com/products/rstudio/download/#download). Both softwares are free. We will provide tutorials on R/RStudio installation and they are also accessible [here](.slides/tutorial.html).

### Resources for R and RStudio

While we will teach you how to effectively use R and RStudio to conduct analyses, one of the key skills required to use R is the ability to find answers on your own. Many common questions or problems are either available on blogs or have been asked and answered in discussion forums already. Finding and deciphering those answers is an important skill you should seek to hone. *You will never remember all of the programming commands!*

Here are some sites where you can find the answers to many R questions and learn new tricks:

 - [_YaRrr, the Pirate's Guide to R_](https://bookdown.org/ndphillips/YaRrr/) by Nathaniel Phillips
 - [_R for Data Science_](https://r4ds.had.co.nz/) by Hadley Wickham
 - [_R Cookbook_](http://www.cookbook-r.com/) by Winston Chang
 - [_An Introduction to Statistical Learning_](http://www-bcf.usc.edu/~gareth/ISL) by Gareth James, Daniela Witten, Trevor Hastie and Robert Tibshirani
 - [Cheat Sheets](https://www.rstudio.com/resources/cheatsheets) 
 - [Quick-R](https://www.statmethods.net) 
 - [StackOverflow](https://stackoverflow.com/questions/tagged/r)
 - [Stack Exchange](https://stats.stackexchange.com)

# Schedule

For more details, see [here](./schedule.html). 

```{r, echo=FALSE, results='asis'}
d <- readxl::read_xlsx(here::here("./data", "syllabus645.xlsx"))
d %>% 
  knitr::kable()
```


# Grading Components and Criteria 
Final grades will be based on the following components:

- Quizzes: 10% (5 quizzes of 2% each)
- Assignments: 60% (5 assignments of 12% each)
- Final: 30%

## Quizzes
Five classes will begin with a very short multiple-choice quiz that is designed to test your knowledge of the theoretical principles underlying the statistics we are studying for the week. While some may feel that this is overly paternalistic, research evidence shows that frequent quizzing increases learning (see a summary of [one study](https://www.nytimes.com/2013/11/21/education/frequent-tests-can-enhance-college-learning-study-finds.html) from the University of Texas). Quizzes will be open book and notes (but not laptop/computer).

## Assignmnets
The goal of the assignments is to practice the concepts and vocabulary we have been modeling in class and implement some of the techniques we have learned. Each assignment has an associated data source, short codebook and detailed instructions for the required data and analytic tasks. You may work on your own or collaborate with one (1) partner. Please make sure that you engage in a a full, fair and mutually-agreeable collaboration if you do choose to collaborate. If you do collaborate, you should plan, execute and write-up your analyses together, not simply divide the work. Please make sure to indicate clearly when your work is joint and any other individual or resource (outside of class material) you consulted in your response. Further assignment details are available [here](./assignments.html).

## Final

# Student Engagement Inventory
Graduate students are expected to perform work of high quality and quantity, typically with forty hours of student engagement for each student credit hour. For this course, the following table shows the number of hours a typical student would expect to spend in each of the following activities:

| Educational activity     | Hours         |    Explanatory comments
|--------------------------|---------------|---------------------------------
| Class attendance         | 30            | TBD frequency
| Class reading and prep   | 15            | Includes reading and review of slides
| Homework Assignments     | 50            | Homework assignments will take 10 hours each (on avg.)
| Final                    | 25            | Includes familiarization with data, data analysis, preparation of displays and writing
| **Total hours**          | **120**       | **These are approximations.** Reading and *especially* analytic time will vary per individual


## Indigenous Recognition Statement
## Role of the GEs  
## Attendance and Absence Guidelines  
## Diversity, Equity, and Inclusion  
## Using Pronouns and Personal Preference  
## Accessible Education
## Reporting Title IX Experiences
## Reporting Obligations
## Academic Misconduct
## Inclement Weather
## Mental Health and Wellness
## Basic Needs
## Accommodation for Religious Observances
## Conflict Resolution
## Grievance Policy
## Course Incomplete Policy
## More per university requirements...



