---
title: "Week 3 Lab"
author: "(add your name)"
date: "4/16/2023"
output:
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
# Don't make a change here
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

Instruction to work on this sheet: To evaluate the level of missingness in the `galo_miss` dataset run the code from steps 1-4, then answer subsequent exercise question. The dataset is available on the class website and can be accessed here: <https://uo-educ-quant.github.io/645/data/galo.csv>

GALO is a 1959 dataset that includes 1,382 sixth grade students in Groningen, Netherlands. The codebook is as follow -

-   schid: School identification code
-   female: an indicator for whether a student is female or now (0: male, 1: female)
-   score: raw score on IQ test (original range 53 to 143)
-   advice: schools advice to students (ordered categorical - coded 0 to 6; one of the 7 groups: No further education; Manual, including housekeeping; Agricultural; Extended primary education; General; Secondary school for girls; Pre-University.)
-   medu: mother education (coded 1 to 9 representing some ordered groups)
-   fedu: mother education (coded 1 to 9 representing some ordered groups)
-   focc: father's occupation (coded 1 to 6; one of the 6 groups: Lower white collar; Middle white collar; Professional, Managers; Shopkeepers; Schooled labor; Unskilled labor)

We might ask: What is the relationship between a teacher's recommendation on secondary school choice and student's IQ score? We are considering the advice variable as continuous for now.

## Topic: Missing Data

### Data

```{r}
#load libraries
library(tidyverse)
library(modelsummary)

## install some new packages if you already don't have these

# install.packages("missMethods")
# install.packages("mice")
library(missMethods)
library(mice)
```

```{r}
#load data from url
load(url("https://uo-educ-quant.github.io/645/data/galo_miss.RData"))

# The data is called galo_miss and shoud have popped up in the environment
head(galo_miss)

#let's re-factor some variables
galo_miss$female <- factor(galo_miss$female,
                         labels = c("Male", "Female"))
```

## 1. Evaluate levels of missingness

```{r}
# two-ways
summary(galo_miss)

# Or use the missMethods package
missMethods::count_NA(galo_miss)
```

### 1.Q. Check variables for missingness and describe in 1-2 sentences.

*Answer Here*

## 2. Plot the patterns and distributions of missing data in the `galo_miss` dataset.

```{r}
# Using the mice package
galo_miss %>% 
  select(advice, medu, fedu, focc) %>% 
  md.pattern(rotate.names = TRUE)
```

### 2.Q. Inspect missingness patterns. Fill this table as a start.

a\. In how many observations is -

-   0 variables missing:

-   1 variable missing:

-   2 variables missing:

-   3 variables missing:

-   4 variables missing:

-   advice missing:

-   medu missing:

-   fedu missing:

-   focc missing:

b\. What's the top three common patterns of missingness?

c\. What is your hunch on the type of missingness for the top three patterns. Is is a) completely at random; b) at random; c) or not random?

## 3. Impute missing data using mice and the cart imputation method

```{r, results  = 'hide'}
# the first argument is the dataset,
# method tells which method of imputation to use: pmm or cart are common
# m tells how many imputed dataset to create: atleast 20 is good
# maxit is the number of iterations: at least 10 is good

galo_multi_cart <- mice(galo_miss, method = "cart", m = 20, maxit = 10)

# Adhoc: try to impute using pmm algorithm and see how different are the results, i.e., create a galo_multi_pmm
```

## 4. Visualize multiple imputation graphs

```{r}
densityplot(galo_multi_cart)
```

### 4.Q. The blue line shows the density plot of the observed data, and the red lines show the density plot of of variables in the imputed dataset. How well was the data imputed, aka, does the distributions match well? Do you think the mechanism of missingness you posited in Q3 is plausible or there is something else going on?

*Answer Here*

## 5. Compare estimates with listwide deletion and mice data

```{r}
#bivariate model with listwise deletion
lm_unconditional_listwise <- lm(advice ~ score, galo_miss)

#multivariate model with listwise deletion
lm_covariate_listwise <- lm(advice ~ score + female + medu + fedu + focc, galo_miss)

#bivariate model with multiply imputed dataset
mice_unconditional_listwise <- with(galo_multi_cart, lm(advice ~ score)) %>%
  pool()

#multivariate model with multiply imputed dataset
mice_covariate_listwise <- with(galo_multi_cart, lm(advice ~ score + 
                              female + medu + fedu + focc)) %>%
  pool()

# Table
modelsummary(list("Bivariate listwise" = lm_unconditional_listwise,
                  "Multivariate listwise" = lm_covariate_listwise,
                  "Bivariate MICE" = mice_unconditional_listwise,
                  "Multivariate MICE" = mice_covariate_listwise),
             stars = TRUE,
             title = "...",
             notes = "...",
             gof_omit = "R2 Adj.|AIC|BIC|Log.Lik.|RMSE",
             coef_rename = c("score" = "IQ score") #can rename other things here too
             )

```

### 5.Q. How does inference change across different methods? What are some tradeoffs here?
