---
title: "Week 2 Lab"
author: "(add your name)"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}

# Set default chunk options: show code, suppress warnings/messages

knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

## *Goals:*


1. Visualize dichotomous outcome: count, proportion, odds ratio
2. Visualize dichotomous outcome and dichotomous predictor

*Instruction:* Each code chunk may be incomplete. Uncomment and complete the code where needed.

### Topic: Understanding Dichotomous and Count Outcomes - Probability and Odds Ratio

### Packages and Data 

```{r}
# Load tidyverse for data manipulation and ggplot2 for visualization
library(tidyverse)

# Load modelsummary for quick summaries of model results
library(modelsummary)

# Read in dataset and select variables of interest
# Make sure the path and column names match your data

prin <- read.csv("https://raw.githubusercontent.com/uo-educ-quant/645/master/data/principal.csv") %>% 
  drop_na() #this is a simplistic way of dealing with missing data, we simply drop observations with missingness on any row
```

For this lab, we will be using the prinicpal.csv dataset available [here](https://raw.githubusercontent.com/uo-educ-quant/645/master/data/principal.csv). 

*Brief variable descriptions:*

 - *schid*, identifier variable for School ID
 
 - *stuid*, identifier variable for Student ID
 
 - *pfemale*, an indicator variable telling if the school principal is a female or not (1 - is a female, 0 - not a female)
 
 - *page*, age of the principal
 
 - *sfemale*, an indicator variable telling if the student is a female or not (1 - is a female, 0 - not a female)
 
 - *match*, an indicator variable telling if the student and principal self-reported sex matched or not (1 - matched, 0 - didn't match)
 
 - *attention*, standardized attention score for the student
 
 - *affective*, an indicator variable telling if the student considers the principal to be affective (1 - Yes, - 0 No)

```{r}
#inspect data
datasummary_skim(prin)
```
**Q1: How many complete observation do we have in the dataset?**
*Answer here*

```{r}
#factor all categorical variables - match, affective, pfemale, sfemale using the code book above
# one is done for you.
 prin$match <- factor(prin$match,
                    levels = c(0,1),
                    labels = c("different", "same"))

  prin$... <- factor(prin$...,
                    levels = c(0,1),
                    labels = c("...", "..."))

  prin$... <- factor(prin$...,
                           levels = c(0,1),
                           labels = c("...", "..."))
  
  prin$... <- factor(prin$...,
                           levels = c(0,1),
                           labels = c("...", "..."))

```


**Table 1. Analytic sample summary statistics**

```{r}
# datasummary formula - specify name of columns ~ statistics for continuous variables

# create a copy of the dataframe where we retain the variables we want to show in the summary

desc_df <- prin %>%
    # select keeps the variable of interest
    select(page, attention, match, sfemale, pfemale, affective)

datasummary_skim(desc_df,
                 # tells which numeric stats to display
                 fun_numeric = list(Mean = Mean, SD = SD, Median = Median),
                 title = "...",
                 notes = "...")
```

**Q2. What proportion of students are labeled as female? What proportion of principles are labeled as female?**


## Dichotomous Outcome

**Q3: Our main outcome variable here is affective.** 


a. How many students consider their principal to be affective? 

b. What is the probability that a student would would say that the principal is affective? (Hint:Students who seen principal as affective /All students )

c. What is the probability that a student would would say that the principal is non-affective? (Hint:Students who seen principal as non affective /All students )

d. What are the odds that students would consider their principal non-affective, compared to being affective?** (Hint: probability not-affective / probability of affective: plug from b and c). 

e. Interpret d in a simple sentence.


```{r}
# Do calculations here to write the above answer. Note that R can be used as a calculator using operators like +,-,*,/,^.
```

### Visualize dichotomous data

**Visualize the number of principals seen as affective vs. not.**

```{r}
# Simple visualization for dichotmous outcome variable - count, proportion

affective_count <- ggplot(data = prin, aes(x = affective))+ # we specify which variables to plot under aes
    geom_bar()+
    labs(x = "...",
         y = "...",
         title = "...")

affective_prop <- ggplot(prin, aes(x = affective))+
    geom_bar(aes(y = after_stat(count/sum(count))))+ # this means calculate proportion before plotting it.
    labs(x = "...",
         y = "...",
         title = "...")
```


**Q3. Present a table and visual displaying the proportion of principals that students to be affective (affective) and the proportion of principals that are female (pfemale). Describe what you see in 2 lines.**

```{r}
# simple contingency table
# outcome 
# datasummary is a flexible function that can make tables based on a formula. It is pretty powerful but will need some investment on your part if you want to understand it better.
# The formula used the convention row ~ column. 
# Which variables should be the rows and which variables should be the column - then, with one side, we can specify which stats should appear using the * sign
datasummary(affective ~ sfemale*(N + Percent("row")), prin,
            title = "...",
            notes = "...",
            fmt = 0)

# Alternate code
table(prin$affective, prin$sfemale,
      dnn = c("Affective", "Gender"))

# To present a table in proportions
# Using dnn labels the table
round(prop.table(table(prin$affective, prin$sfemale,
      dnn = c("Affective", "Female")),2),2)
```

**Visualizing dichotmous outcome and predictor**

```{r}
# Can show counts - okay representation
# Can show counts - okay representation
# With binary outcomes and predictors, the predictor goes on x axis and outcomes goes on fill
ggplot(prin, aes(x = sfemale, fill = affective))+
    geom_bar(position = "dodge")+
    labs(x = "...",
         y = "...",
         title = "...")


# Can show proportions, which is way better
ggplot(prin, aes(x = sfemale, fill = affective))+
    geom_bar(position = "fill")+
    labs(x = "...",
         y = "...",
         title = "...")
```


**Q5a. What are the odds that students would consider their principal non-affective, compared to being affective for female principals? (see results from table to answer)** 

**Q5b. What are the odds that students would consider their principal non-affective, compared to being affective for male principals?** 

**Q5c. What do you descriptively think about the relationship between principal's gender and students perception of affectiveness? (in 1-2 lines)**
