---
title: "Week 5 Lab"
author: "(add your name)"
date: "4/30/2023"
output:
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}

# Don't make a change here
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

## *Goals:*

1. Visualizing dichotomous outcomes with continuous and categorical explanatory variables,
2. Conduct a simple logistic regression with continuous predictor. 
3. Conduct a simple logistic regression with categorical predictor. 

*Instruction to work on this sheet:* In each code chunk, you can find some code being taken out purposely and left with three dots instead. Your task is to first uncomment the code line, replace the three dots with the code you would write, and then run the code to proceed.

*Data*
For this lab, we will be using the prinicpal.csv dataset available [here](https://raw.githubusercontent.com/uo-educ-quant/645/master/data/principal.csv). 
Read in the dataset in the environment and save it to the object prin.

*Brief variable descriptions:*

 - *schid*, identifier variable for School ID
 
 - *stuid*, identifier variable for Student ID
 
 - *pfemale*, an indicator variable telling if the school principal is a female or not (1 - is a female, 0 - not a female)
 
 - *page*, age of the principal
 
 - *sfemale*, an indicator variable telling if the student is a female or not (1 - is a female, 0 - not a female)
 
 - *match*, an indicator variable telling if the student and principal self-reported sex matched or not (1 - matched, 0 - didn't match)
 
 - *attention*, standardized attention score for the student
 
 - *affective*, an indicator variable telling if the student considers the principal to be affective (1 - Yes, - 0 No)

### Topic: Dichotomous outcomes and basic logistic regression

### Packages and Data 

```{r}
# Load necessary packages 
library(tidyverse)
library(modelsummary)
#install.packages("broom")
library(broom)

# read in dataset and select variables of interest 
# Quotes are important around URL link for data retrieval 

prin <- read.csv("https://raw.githubusercontent.com/uo-educ-quant/645/master/data/principal.csv") %>% 
  drop_na() #this is a simplistic way of dealing with missing data, we simply drop observations with missingness on any row
```

**RQ1: What is the relationship between principal age _page_ and student perception of principal affectiveness _affective_ (1 - Yes, - 0 No)?**

**Task 1: Prepare your dataset for analysis by selecting the relevant variables, and transforming to factor wherever needed.**

```{r}
# check that variables are the correct types. 
# Dichotomous variables should be factors, and continuous variables should be numeric (num) or integers (int).
# factor all categorical variable


#str(prin)

# convert any dichotomous variable to factor
prin$pfemale <- factor(prin$pfemale,
                    levels = c(0,1),
                    labels = c("Male", "Female"))
 
prin$sfemale <- factor(prin$sfemale,
                    levels = c(0,1),
                    labels = c("Male", "Female"))
  
prin$affective <- factor(prin$affective, 
                           levels = c(0,1),
                           labels = c("Not Affective", "Affective"))

prin$match <- factor(prin$match, 
                           levels = c(0,1),
                           labels = c("Different", "Same"))
```


```{r}
# Descriptive table
prin_desc <- prin %>%
    select(-c(schid, stuid)) %>%
    rename(
        "Principal Gender" = pfemale,
        "Student Gender" = sfemale,
        "Principal-Student Gender Match" = match,
        "Principal Age" = page,
        "Student Std. Attention Score" = attention,
        "Student's perception of Principal Affectiveness" = affective
    )

datasummary_skim(prin_desc,
                 fun_numeric = list(Mean = Mean, SD = SD))
```

### Logistic regression model estimates (continuous IV)

RQ: What is the relationship between principal's age and student's perception of principal affectiveness?

**Q: Write the independent and dependent variables? What is the null hypothesis?**


```{r}
# Some raw data visualization
# The main predictor variable should be on the x-axis and outcome should have other aesthetic - like fill. 

#Method 1
ggplot(prin, aes(x = page, fill = affective)) +
    geom_density(alpha = 0.4)+
    theme_minimal() +
    labs(x = "Principal Age",
         y = "Distribution",
         fill = "Student perception")

#Method 2
ggplot(prin, aes(x = page, fill = affective)) +
    geom_density(alpha = 0.4, show.legend = FALSE)+
    theme_minimal() +
    facet_wrap(~affective)+
    labs(x = "Principal Age",
         y = "Distribution")
    
```

```{r}
# family refers to the type of distribution, where we have specified binomial, and the link asks for the link function that we want to use, which is the logit for a logistic regression.
# The formula has to be of the form y ~ x

# mod_1 <- glm(... ~ ..., data = ...,
#              family = binomial(link = "logit"))

# this gives the output in log odds unit
summary(mod_1)
```

In a multiple logistic regression model, which includes two or more predictor variables, exponentiation of the model parameters is typically necessary for interpretation. This is because the coefficients in multiple logistic regression represent the change in the log odds of the outcome variable for a one-unit change in the predictor variable, holding all other variables constant. Exponentiating the coefficients results in odds ratios, which provide a more interpretable measure of effect size.

```{r}
# this gives the output in odds ration units

# mod_1 %>%
#       broom::tidy(exponentiate = TRUE, conf.int = TRUE) 
        
```

**Interpret the result.**

**Q: What is the estimate of the relationship? What is the direction of the estimated relationship?**

**Q: Is the estimate significant at the 0.05 alpha threshold?**

**Q: Interpret the relationship.**

### Logistic regression model estimates (categorical IV)

RQ2: Do female principals have greater odds of being considered affective than male principals?

**Q: Write the independent and dependent variables? What is the null hypothesis?**

```{r}
# plot
ggplot(prin, aes(x = pfemale, fill = affective)) +
    geom_bar(position = "dodge")+
    theme_minimal() +
    labs(x = "Principal Gender",
         y = "Number of Students",
         fill = "Student perception")
```


```{r}
# mod_2 <- glm(y ~ x, data = ...,
#              family = binomial(link = "logit"))
# 
# summary(mod_2)

# mod_2 %>%
#       broom::tidy(exponentiate = TRUE, conf.int = TRUE) 
    
```

**Interpret the result.**

**Q: What is the estimate of the relationship? What is the direction of the estimated relationship?**

**Q: Is the estimate significant at the 0.05 alpha threshold?**

**Q: Interpret the relationship.**

## Optional Part

### Logistic regression model estimates (categorical and continuous IV)

**RQ3: Do female principals have greater odds of being considered affective than male principals, after adjusting for principal age?**

**Q: Write the independent and dependent variables? What is the null hypothesis?**


```{r}
# mod_3 <- glm(y ~ x + covariate, data =...,
#              family = binomial(link = "logit"))
# 
# summary(mod_3)
```


```{r}
# mod_3 %>% 
#       broom::tidy(exponentiate = TRUE, conf.int = TRUE)

```

**Q: What is the estimate of the relationship? What is the direction of the estimated relationship?**

**Q: Is the estimate significant at the 0.05 alpha threshold?**

**Q: Interpret the relationship.**

**Q: How is this different from the previous two parts?**





