---
title: "Week 8 Lab"
author: "(add your name)"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}

# Don't make a change here
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      eval = TRUE)
```

*Goals:*

1. Conduct a mixed-effect logistic regression in nested data: Random intercepts and slopes 
2. Model comparison, model of best fit
3. Diagnostics

*Instruction to work on this sheet:* In each code chunk, you can find some code being taken out purposely and left with three dots instead. Your task is to first uncomment the code line, replace the three dots with the code you would write, and then run the code to proceed.

*Data*
For this lab, we will be using the prinicpal.csv dataset available [here](https://raw.githubusercontent.com/uo-educ-quant/645/master/data/principal.csv). 
Read in the dataset in the environment and save it to the object prin.

*Brief variable descriptions:*

 - *schid*, identifier variable for School ID
 
 - *stuid*, identifier variable for Student ID
 
 - *pfemale*, an indicator variable telling if the school principal is a female or not (1 - is a female, 0 - not a female)
 
 - *page*, age of the principal
 
 - *sfemale*, an indicator variable telling if the student is a female or not (1 - is a female, 0 - not a female)
 
 - *match*, an indicator variable telling if the student and principal self-reported sex matched or not (1 - matched, 0 - didn't match)
 
 - *attention*, standardized attention score for the student
 
 - *affective*, an indicator variable telling if the student considers the principal to be affective (1 - Yes, - 0 No)
 
 
 
### Topic: Mixed-effects Logistic Regression in nested data

### Packages and Data 

```{r}
# Load necessary packages 
library(tidyverse)
#install.packages("broom")
library(broom)
#install.packages("modelsumary")
library(modelsummary)
#install.packages("lme4")
library(lme4)
# read in dataset and select variables of interest 
# Quotes are important around URL link for data retrieval 

prin <- read.csv("https://raw.githubusercontent.com/uo-educ-quant/645/master/data/principal.csv") %>% 
  drop_na() #this is a simplistic way of dealing with missing data, we simply drop observations with missingness on any row
```

## Data Prepration

**Prepare your dataset for analysis by selecting the relevant variables, and transforming to factor wherever needed. We're also going to be factoring school IDs to nest.**

In educational research, students are typically grouped within classes and schools, creating a hierarchical data structure. This nesting results in students within the same school not being entirely independent, challenging the assumption of independence required by many statistical models. Schools vary in resources, teaching methods, demographics, and policies, all of which can affect study outcomes. Including school ID as a random effect in your statistical model helps account for these differences between schools, leading to a more accurate and reliable analysis of the data at the individual or classroom level.

```{r}
# check that variables are the correct types. 
# Dichotomous variables should be factors, and continuous variables should be numeric (num) or integers (int).
# factor all categorical variable

# convert any dichotomous variable to factor
prin$pfemale <- factor(prin$pfemale,
                    levels = c(0,1),
                    labels = c("Male", "Female"))
 
prin$sfemale <- factor(prin$sfemale,
                    levels = c(0,1),
                    labels = c("Male", "Female"))
  
prin$affective <- factor(prin$affective,
                             levels = c(0, 1), 
                          labels = c("No", "Yes"))
 
prin$match <- factor(prin$match, 
                           levels = c(0,1),
                           labels = c("Didn't match", "Matched"))
  
prin$schid <- as.factor(prin$schid)

str(prin)
```

**RQ1: What is the relationship between student-principal gender matching and student perceptions of principal effectiveness?**

**1. Write the outcome, predictor, and nesting variable?**

### Visualize the nested structure

```{r}
# let's see how many students per school we have in the sample
table(prin$schid)

# let's visualize this nested structure

ggplot(prin, aes(y = schid))+
  geom_bar(fill = "cornflowerblue")+
  labs(x = "School ID",
       y = "Number of students per school") +
  theme_minimal()

ggplot(prin, aes(x = affective, 
                 y = schid, color = affective))+
  geom_jitter()+
  theme_minimal()
```


**2. What are your observations about the data structure? How many students per school? Are there schools where more or less people think of the principal as affective?**

## Accounting for Nested Variable

**Compare null model (no predictors) to model with predictor, both including random intercepts**

**Next, we will answer this question: what is the between-school variance in the outcome?** 


```{r}
### Random Intercepts Mixed-effects Model ###

# By specifying “between-school variance,” the question implicitly suggests examining how the intercepts (i.e., the baseline levels of perceived principal effectiveness) vary from one school to another.

rq1_null <- glmer(outcome ~ 1 + (1|nesting_variable), data =prin,
             family = binomial(link = "logit"))

# Access ICCs between null 

performance::icc(rq1_null)
```

**3. What's the between-school variance in the affectiveness variable?**

**4. Is there reason to believe there is between school clustering and should we account for this in our model?**


```{r}
rq1_mod1 <- glmer(outcome ~ predictor + (1|nesting_variable), data =prin,
             family = binomial(link = "logit"))

performance::icc(rq1_mod1)
```


```{r}
# running a model not accounting for clustering
rq1_wo_nesting <- glm(affective ~ match, data =prin,
             family = binomial(link = "logit"))
```


```{r}
## A neater table        
modelsummary(list('Without Clustering' = rq1_wo_nesting, 
                  'Mixed-Effect Model' = rq1_mod1),
             exponentiate = TRUE,
             statistic = "conf.int",
             stars = TRUE,
             gof_omit = "AIC|BIC|Log.Lik|RMSE",
             title = "Relationship between ....",
             notes = "Data ....")
```

**5. What changed between the two models when we didn't account and then accounted for clustering?**

## Random Slopes and Model Comparison

**RQ2: How does the relationship between student-principal gender matching and student perceptions of principal effectiveness vary across schools, and to what extent do schools differ in this relationship?**


```{r}
### Random Intercepts and Slopes Mixed-effects Model ###

rq2_mod2 <- glmer(outcome ~ predictor + (1 + predictor |nesting_variable), data =prin,
             family = binomial(link = "logit"))

# compare random intercepts to random intercepts and slopes model (i.e., mod1 to mod2)

performance::compare_performance(rq1_null, rq1_mod1, rq2_mod2, metrics = "common", verbose = FALSE) %>% 
  datasummary_df()


performance::test_likelihoodratio(rq1_null, rq1_mod1, rq2_mod2) %>% 
  datasummary_df()
```

**6. Which is the better fitting model?**

```{r}
## A neater table        
modelsummary(list(
    'Null Model' = rq1_null,
    'Random Intercept' = rq1_mod1, 
    'Random Slopes' = rq2_mod2),
             exponentiate = TRUE,
             statistic = "conf.int",
             stars = TRUE,
             gof_omit = "AIC|BIC|Log.Lik|RMSE",
             title = "Relationship between ....",
             notes = "Data ....")
```

**7. Interpret the best fitting model.**